---
title: "trantract: extract and visualize transit information by census tract"
author: "Carl Parker"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: styles.css
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# trantract: extract and visualize transit information by census tract #

**NB**: This package requires a dev version of `ggmap()` to avoid to avoid
the following error:

    Error: GeomRasterAnn was built with an incompatible version of
    ggproto.  Please reinstall the package that provides this extension.

You can use the following command to install the dev version of
`ggmap()`:

    install_github("dkahle/ggmap")

See the following URL for more information:

  <http://www.arilamstein.com/blog/2017/01/16/choroplethr-v3-5-3-now-cran/>

---

The trantract package enables you to explore and visualize how King
Country Metro transit service is related to the spatial characteristics
of US Census Bureau census tracts. The package also includes a data set
of some demographic information associated with census tracts that can
be explored as well.

In this vignette, we'll use the functionality and data in the package to
create a measure of the level of transit service ("route-stops") and see
how that measure varies across census tracts. We'll then perform some
simple modeling to see if there is any clear relationship between level
of service and certain demographic variables.

Load the library


```{r warning = FALSE, message = FALSE, echo = TRUE}
library( trantract )
```

Load the data sets

```{r}
data( stops )
data( kct.shp.proj )
data( route.stops.df )
data( tract.demographics.kc )
```

Get a list of all census tracts.
 
```{r}
census.tracts <- unique( stops$tract )
```

For each tract, get the count of how many route-stops there are in the
tract. For example, if a stop is served by four routes, that equates
to four route-stops.

Basically, we are creating a measure of how well King County metro
serves that tract.

Create a DF that initially has one column: the census tracts.

```{r}
service.distro.df <- data.frame( tract = census.tracts )

route.stops.vec <- numeric( 0 )
```

For each tract, determine the number of route-stops.

```{r}
for( tract in service.distro.df$tract ) {

  route.stops.vec[ tract ] <- nrow( route_stops_from_census_tract( tract ) )

}
```

Add the route-stops as a column to our service DF

```{r}
service.distro.df$route.stops <- route.stops.vec
```

See how the level of service is distributed.

```{r}
hist( service.distro.df$route.stops ) 
```

That is pretty skewed. Try a log transform.

```{r}
hist( log( service.distro.df$route.stops ) ) 
```

Get the min and max and look at those census tracts

```{r}
service.distro.df[ service.distro.df$route.stops == min( service.distro.df$route.stops ), ]
service.distro.df[ service.distro.df$route.stops == max( service.distro.df$route.stops ), ]

plot_tract( 320.07, map.zoom = 13 )
plot_tract( 81, map.zoom = 15 )
```

Now let's look at how our measure of service relates to some demographic
information.

The Census.Tract column in the demographic data is character, not
factor. So create a similar column in our data to facilitate the
merge.

```{r}
service.distro.df$Tract <- as.character( service.distro.df$tract )
tract.demographics.kc.routes <- merge( tract.demographics.kc, service.distro.df )  
```

First, we'll regress against: total population of the tract, total
housing units, and total land area.

```{r}
mod.tract.multivar <- lm( route.stops ~ Population + Housing.Units + Land.Area, data = tract.demographics.kc.routes )
summary( mod.tract.multivar )
```

That is not a very compelling model.

Let's just regress against the values for density in our demographic
data set.

```{r set-options, warning = FALSE, message = FALSE, echo = FALSE}
options( "width" = 120 )
mod.tract.density <- lm( route.stops ~ Population.Density, data = tract.demographics.kc.routes )
summary( mod.tract.density )
```

Interestingly, that doesn't give a strong relationship either. 

Visualize it.

```{r}
plot( 
     tract.demographics.kc.routes$Population.Density, 
     log( tract.demographics.kc.routes$route.stops ), 
     main = "Level of service vs density",
     xlab = "Persons / Square Miles",
     ylab = "log( route-stops )"
    )
```

People often say that transit is planned around "urban density" but from
these simple regressions, it appears that there must other factors as
well.

Let's keep going.

Change the response variables so that we aren't just looking at number
of route-stops, but the density of route-stops; not all census tracts
are the same size. 

Add a column that represents the _density_ of the route.stops in each
tract. Note, again, that stops in and of themselves are _not_ a good
measure of service because of the wide variation in the number of
buses that serve any given stop. (And even this is a simplification
because we aren't considering number of trips that any given bus makes
to that stop each day--what most people would think of as frequency.)

```{r}
tract.demographics.kc.routes$route.stops.dens <- tract.demographics.kc.routes$route.stops / tract.demographics.kc.routes$Land.Area
```


Let's also use the lat-lon, but we will normalize these.

```{r}
mean.INTPTLAT10 <- mean( tract.demographics.kc.routes$INTPTLAT10 )
sd.INTPTLAT10   <- sd(   tract.demographics.kc.routes$INTPTLAT10 )

mean.INTPTLON10 <- mean( tract.demographics.kc.routes$INTPTLON10 )
sd.INTPTLON10   <- sd(   tract.demographics.kc.routes$INTPTLON10 )

norm_lat <- function( lat.arg ) {

  ( lat.arg - mean.INTPTLAT10 ) / sd.INTPTLAT10 

}

norm_lon <- function( lon.arg ) {

  ( lon.arg - mean.INTPTLON10 ) / sd.INTPTLON10 

}

tract.demographics.kc.routes$norm.intptlat10 <- norm_lat( tract.demographics.kc.routes$INTPTLAT10 )
tract.demographics.kc.routes$norm.intptlon10 <- norm_lon( tract.demographics.kc.routes$INTPTLON10 )
```

How many tracts are "north" and "south" of zero.

```{r}
sum( tract.demographics.kc.routes$norm.intptlat10 < 0 ) 
sum( tract.demographics.kc.routes$norm.intptlat10 > 0 ) 
```


We've set the stage, let's rock a few more regressions.


First, let's try route.stops.dens (density of route.stops) as the response.

```{r}
mod.tract.rs.dens.density <- lm( route.stops.dens ~ Population.Density, data = tract.demographics.kc.routes )
summary( mod.tract.rs.dens.density )
```

Wow. Okay. That made a difference! The R-Squared is no 0.24.

Try quadratic and cubic functions of the population density.

```{r}
mod.tract.rs.dens.density.quad <- lm( route.stops.dens ~ poly( Population.Density, 2, raw = TRUE ), data = tract.demographics.kc.routes )
summary( mod.tract.rs.dens.density.quad )

mod.tract.rs.dens.density.cube <- lm( route.stops.dens ~ poly( Population.Density, 3, raw = TRUE ), data = tract.demographics.kc.routes )
summary( mod.tract.rs.dens.density.cube )
```

How do these three compare when measured against `AIC()`?

```{r}
AIC( mod.tract.rs.dens.density, mod.tract.rs.dens.density.quad, mod.tract.rs.dens.density.cube )
```

In StatR 501 they said that a difference in AIC > 10 was strong evidence
that a model was better. On that basis, we would choose the cubic model.

Look at some visualizations.

```{r}
plot( 
     tract.demographics.kc.routes$Population.Density, 
     tract.demographics.kc.routes$route.stops.dens, 
     main = "Level of service (density) vs population density",
     xlab = "Persons / Square Mile",
     ylab = "route-stops-density"
    )

intercept <- coef( mod.tract.rs.dens.density )[ "(Intercept)" ]
slope     <- coef( mod.tract.rs.dens.density )[ "Population.Density" ]

abline( intercept, slope )
```

Show the population density that gets the greatest density of route
stops.

```{r}
( max.rs.dens <- max( tract.demographics.kc.routes$route.stops.dens ) ) 
( pop.dens.for.max.service <- tract.demographics.kc.routes[ tract.demographics.kc.routes$route.stops.dens == max.rs.dens, ]$Population.Density )
abline( v = pop.dens.for.max.service, col = "red" )
```

Show the route stop density that is given to the tract with the
greatest population density.

```{r}
( max.pop.dens <- max( tract.demographics.kc.routes$Population.Density ) )
( rs.dens.for.max.pop.dens <- tract.demographics.kc.routes[ tract.demographics.kc.routes$Population.Density == max.pop.dens, ]$route.stops.dens )
abline( h = rs.dens.for.max.pop.dens, col = "green" )


( cube.coef <- coef( mod.tract.rs.dens.density.cube ) )

x.cube <- 0:50000
y.cube <- vapply( 0:50000, function( x ) y = cube.coef[ 4 ] * ( x ** 3 ) + cube.coef[ 3 ] * ( x ** 2 ) + cube.coef[ 2 ] * ( x ) + cube.coef [ 1 ], numeric( 1 ) )
points( x.cube, y.cube, cex = 0.10, col = "blue" ) 
```


Use log() scale for the y-axis.

```{r}
plot( 
     tract.demographics.kc.routes$Population.Density, 
     log( tract.demographics.kc.routes$route.stops.dens ), 
     main = "Level of service (density) vs population density",
     xlab = "Persons / Square Mile",
     ylab = "log( route-stops-density )"
    )

#
# Show the population density that gets the greatest density of route
# stops.
#
```{r}
abline( v = pop.dens.for.max.service, col = "red" )
```

Show the route stop density that is given to the tract with the
greatest population density.

```{r}
( log.rs.dens.for.max.pop.dens <- log( rs.dens.for.max.pop.dens ) )
abline( h = log.rs.dens.for.max.pop.dens, col = "green" )

log.y.cube <- log( y.cube )
points( x.cube, log.y.cube, cex = 0.10, col = "blue" ) 
```

Now, regress against the normalized lat and lon.

```{r}
mod.tract.latlon <- lm( route.stops.dens ~ norm.intptlat10 + norm.intptlon10, data = tract.demographics.kc.routes )
summary( mod.tract.latlon )
```

What is funny is that you can see that transit service falls off rapidly
as you head west...into the ocean.


As with population density, let's try polynomial functions of the
covariates.

```{r}
mod.tract.latlon.quad <- lm( route.stops.dens ~ poly( norm.intptlat10, 2, raw = TRUE ) + poly( norm.intptlon10, 2, raw = TRUE ), data = tract.demographics.kc.routes )
summary( mod.tract.latlon.quad )

mod.tract.latlon.cube <- lm( route.stops.dens ~ poly( norm.intptlat10, 3, raw = TRUE ) + poly( norm.intptlon10, 3, raw = TRUE ), data = tract.demographics.kc.routes )
summary( mod.tract.latlon.cube )
```

And here are the `AIC()`.

```{r}
AIC( mod.tract.latlon, mod.tract.latlon.quad, mod.tract.latlon.cube )
```

Based on the AIC(), the cubic model is outperforming the other two.

Visualize the distribution of service across the lat-lon values.

Give each point one of three colors to represent its service level.

```{r}
log.rs.dens.5n  <- fivenum( log( tract.demographics.kc.routes$route.stops.dens ) )
tukey.low.hinge <- log.rs.dens.5n[ 2 ]
tukey.up.hinge  <- log.rs.dens.5n[ 4 ]

color.tukey <- function( val ) {
  ifelse( log( val ) < tukey.low.hinge, "red", ifelse( log( val ) > tukey.up.hinge, "green", "blue" ) )
}

tract.demographics.kc.routes$color <- color.tukey( tract.demographics.kc.routes$route.stops.dens )

plot( x = tract.demographics.kc.routes$norm.intptlon10, 
      y = tract.demographics.kc.routes$norm.intptlat10,
      col = tract.demographics.kc.routes$color,
      main = "Normalized location of tracts",
      xlab = "Normalized longitude of tract internal point",
      ylab = "Normalized latitude of tract internal point"
    )

abline( h = 0, v = 0, col = "gray", lwd = 2, lty = 1 )

legend( x = "right", pch = 1, legend = c( "High", "Med", "Low" ), col = c( "green", "blue", "red" ), title = "Service level" ) 
```


What is interesting about this graph is that you can see instances where
you have tracts with high service (green) adjacent to tracts with low
service (red). One hopes that this is accounted for by other factors
such as a pocket of low population density.



Put it all together.

```{r}
mod.tract.density.latlon.cube <- lm( route.stops.dens ~ 
                                       poly( Population.Density, 3, raw = TRUE ) +
                                       poly( norm.intptlat10, 3, raw = TRUE ) + poly( norm.intptlon10, 3 ), 
                                       data = tract.demographics.kc.routes 
                                   )

summary( mod.tract.density.latlon.cube )
AIC( mod.tract.density.latlon.cube )
```

When we combined lat-lon with population density, we seem to have lost
all the benefit of lat-lon.  The AIC() is not particularly better than
using population density alone, nor is the R-Squared.  Also, the errors
for the lat-lon terms are high, high.

This makes me think that to the extent that lat-lon was working, it was
working because it somehow reflected population density information.

I think lat-lon is also problematic because Seattle is an isthmus--and
it is a rather strangely shaped isthmus because of Elliott Bay. It also
has two significant inland bodies of water: Lake Union and Green Lake.

The next step in this investigation would be to somehow include service
frequency information for each stop. I did not include that in my
initial investigation because I didn't see any quick and easy way to
extract it from the transit description files published by King County
metro.

