---
title: "trantract: extract and visualize transit information by census tract"
author: "Carl Parker"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: styles.css
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# trantract: extract and visualize transit information by census tract #

**NB**: This package requires a dev version of `ggmap()` to avoid to avoid
the following error:

    Error: GeomRasterAnn was built with an incompatible version of
    ggproto.  Please reinstall the package that provides this extension.

You can use the following command to install the dev version of
`ggmap()`:

    install_github("dkahle/ggmap")

See the following URL for more information:

  <http://www.arilamstein.com/blog/2017/01/16/choroplethr-v3-5-3-now-cran/>

---

The trantract package enables you to explore and visualize how King
Country Metro transit service is related to the spatial characteristics
of US Census Bureau census tracts. The package also includes a data set
of some demographic information associated with census tracts that can
be explored as well.

In this vignette, we'll use the functionality and data in the package to
create a measure of the level of transit service ("route-stops") and see
how that measure varies across census tracts. We'll then perform some
simple modeling to see if there is any clear relationship between level
of service and certain demographic variables.

Load the library


```{r warning = FALSE, message = FALSE, echo = TRUE}
library( trantract )
```

Load the data sets

```{r}
data( stops )
data( kct.shp.proj )
data( route.stops.df )
data( tract.demographics.kc )
```

Get a list of all census tracts.
 
```{r}
census.tracts <- unique( stops$tract )
```

For each tract, get the count of how many route-stops there are in the
tract. For example, if a stop is served by four routes, that equates
to four route-stops.

Basically, we are creating a measure of how well King County metro
serves that tract.

Create a DF that initially has one column: the census tracts.

```{r}
service.distro.df <- data.frame( tract = census.tracts )

route.stops.vec <- numeric( 0 )
```

For each tract, determine the number of route-stops.

```{r}
for( tract in service.distro.df$tract ) {

  route.stops.vec[ tract ] <- nrow( route_stops_from_census_tract( tract ) )

}
```

Add the route-stops as a column to our service DF

```{r}
service.distro.df$route.stops <- route.stops.vec
```

See how the level of service is distributed.

```{r}
hist( service.distro.df$route.stops ) 
```

That is pretty skewed. Try a log transform.

```{r}
hist( log( service.distro.df$route.stops ) ) 
```

Get the min and max and look at those census tracts

```{r}
service.distro.df[ service.distro.df$route.stops == min( service.distro.df$route.stops ), ]
service.distro.df[ service.distro.df$route.stops == max( service.distro.df$route.stops ), ]

plot_tract( 320.07, map.zoom = 13 )
plot_tract( 81, map.zoom = 15 )
```

Now let's look at how our measure of service relates to some demographic
information.

The Census.Tract column in the demographic data is character, not
factor. So create a similar column in our data to facilitate the
merge.

```{r}
service.distro.df$Census.Tract <- as.character( service.distro.df$tract )
tract.demographics.kc.routes <- merge( tract.demographics.kc, service.distro.df )  
```

First, we'll regress against: total population of the tract, total
housing units, and total land area.

```{r}
mod.tract.multivar <- lm( route.stops ~ Total.Population + Total.Housing.Units + Land.Area.Square.Miles, data = tract.demographics.kc.routes )
summary( mod.tract.multivar )
```

That is not a very compelling model.

Let's just regress against the values for density in our demographic
data set.

```{r set-options, warning = FALSE, message = FALSE, echo = FALSE}
options( "width" = 120 )
mod.tract.density <- lm( route.stops ~ Population.Density.Persons.Square.Mile, data = tract.demographics.kc.routes )
summary( mod.tract.density )
```

Interestingly, that doesn't give a strong relationship either. 

Visualize it.

```{r}
plot( 
     tract.demographics.kc.routes$Population.Density.Persons.Square.Mile, 
     log( tract.demographics.kc.routes$route.stops ), 
     main = "Level of service vs density",
     xlab = "Persons / Square Miles",
     ylab = "log( route-stops )"
    )
```

People often say that transit is planned around "urban density" but from
these simple regressions, it appears that there must other factors as
well.

